Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2011-06-30 appendix/appendix.tex
a day of writing

\section{Model Derivations}\label{modelderivations}
The sudden collapses we wish to detect are driven by bifurcations when an eigenvalue changes sign in the corresponding deterministic model,
resulting in the sudden loss of a stable state and the onset of a critical transition.
There are many possible models that can contain such a transition.
These may be highly nonlinear models with multiple attractors.
The diversity of possible models can be classified by the type of bifurcation,
such as the saddle-node bifurcation in which a stable point (node) and unstable point (saddle) collide and annihilate,
or a trans-critical bifurcation, in which a stable point becomes unstable.  
(Any dynamical systems textbook, such as \citep{Guckenheimer1983} can give a good overview of such models and bifurcations). 
In general we will be unable to estimate such models accurately if the system has only been observed near the favorable attractor (i.e. before the critical transition occurs),
even if we knew the functional form of the model.  Fortunately the detection of early warning is based on a \emph{local} theory, predicting changes in the dynamics near the current attractor.
Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model, but only the linearized model near the bifurcation.

>>>>

\section{Model Derivations}\label{modelderivations}

The sudden collapses we wish to detect are driven by bifurcations when an eigenvalue changes sign in the corresponding deterministic model,
resulting in the sudden loss of a stable state and the onset of a critical transition.
There are many possible models that can contain such a transition.
These may be highly nonlinear models with multiple attractors.
The diversity of possible models can be classified by the type of bifurcation,
such as the saddle-node bifurcation in which a stable point (node) and unstable point (saddle) collide and annihilate,
or a trans-critical bifurcation, in which a stable point becomes unstable.  
(Any dynamical systems textbook, such as \citep{Guckenheimer1983} can give a good overview of such models and bifurcations). 
In general we will be unable to estimate such models accurately if the system has only been observed near the favorable attractor (i.e. before the critical transition occurs),
even if we knew the functional form of the model.  Fortunately the detection of early warning is based on a \emph{local} theory, predicting changes in the dynamics near the current attractor.
Consequently, it is as unnecessary as it is impossible to estimate a fully nonlinear model, but only the linearized model near the bifurcation.
